### 2013-08-23 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
spell check for nonlinear chapter

In this chapter the fundamentals of Newton's method will be presented
along with a brief discussion of inexact Newton methods. Newton-Krylov
methods, a variant of inexact Newton methods, will be outlined
including an automatic technique for generating the Jacobian commonly
used in today's applications. Following this, a new nonlinear solution
algorithm utilizing Monte Carlo Synthetic Acceleration will be
presented and implementation details will be discussed. Using the new
algorithm and a production fluid physics code, a group of challenging
benchmark problems for the Navier-Stokes equations will be used to
verify the correctness of the method relative to solutions obtained
using conventional methods in various flow regimes and
geometries. Finally, we will compare the performance of the new method
against contemporary Newton-Krylov methods for the same benchmark
problems.

@@

\subsection{Inexact Newton Methods}
\label{subsec:inexact_newton_methods}
Inexact Newton methods arise when the Jacobian operator is not exactly
inverted, resulting in an inexact Newton correction as initially
described by Dembo and others \cite{dembo_inexact_1982}. For common
sparse nonlinear systems, which in turn yield a sparse Jacobian
matrix, this situation occurs when conventional iterative methods are
applied. In their definition, Dembo formulated inexact methods such
that they are independent of the linear method used to solve for the
Newton correction and therefore are amenable to use with any linear
solver. Furthermore, they bind the convergence of the outer nonlinear
iteration to the inner linear iteration such that:
\begin{equation}
  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
  ||\ve{F}(\ve{u}^k)||\:,
  \label{eq:inexact_newton_forcing}
\end{equation}
where $\eta^k \in [0,1)$ is defined as the \textit{forcing term} at
  the $k$ iterate. Eq~(\ref{eq:inexact_newton_forcing}) then states
  that the residual generated by the linear solver is bound by the
  nonlinear residual and how tightly it is bound is defined by the
  forcing term. This is useful in that we can vary how tightly coupled
  the convergence of the linear iterations used to generate the Newton
  correction is to the nonlinear iteration by relaxing or tightening
  the convergence properties on the linear iterative method. As a
  result, strategies for determining the forcing term can vary
  depending on the problem type and can greatly affect the convergence
  of the method or even prohibit convergence
  \cite{eisenstat_choosing_1996}. In addition, \textit{globalization
    methods} may be used to modify the Newton correction in a more
  desirable direction such that convergence properties can be
  improved when the initial guess for $\ve{u}$ is poor
  \cite{pawlowski_globalization_2006}.

\subsection{Newton-Krylov Methods\ }
\label{subsec:newton_krylov_methods}
A form of inexact Newton methods, \textit{Newton-Krylov methods} are
nonlinear iterative methods that leverage a Krylov subspace method as
the linear solver for generating the Newton correction
\cite{kelley_iterative_1995}. As outlined in
Appendix~\ref{ch:linear_problem}, Krylov methods are robust and enjoy
efficient parallel implementations on modern
architectures. Furthermore, their lack of explicit dependence on the
operator make them easier to implement than other
methods. Additionally, although many iterations can become memory
intensive due to the need to store the Krylov subspace for the
orthogonalization procedure, at each nonlinear iteration this cost is
reset as the Jacobian matrix will change due to its dependence on the
solution vector. This means that for every nonlinear iteration, a
completely new linear system is formed for generating the Newton
correction and we can modify the Krylov solver parameters accordingly
to accommodate this. In most nonlinear problems, the Jacobian operator
is generally non-symmetric and therefore either Krylov methods with
long recurrence relations that can handle non-symmetric systems must
be considered or the Newton correction system must be preconditioned
such that the operator is symmetric and short recurrence relation
methods can be potentially be used.

With many Krylov methods available, which to use with the Newton
method is dependent on many factors including convergence rates and
memory usage. Several studies have been performed to investigate this
\cite{mchugh_inexact_1993,knoll_newton-krylov_1995}. In their
numerical studies in 1995, Knoll and McHugh used the set of highly
nonlinear and stiff convection-diffusion-reaction equations to solve a
set of tokamak plasma problems with the goal of measuring solver
performance with Newton's method. They note several trade-offs in
using Krylov methods with the Newton solver. The first is that the
optimization condition that results from the constraints (e.g. the
minimization of the GMRES residual over the Krylov space) can be
relaxed by restricting the size of the subspace such that only a fixed
number of subspace vectors may be maintained, thus reducing memory
requirements.

We can also relax the optimization condition by instead restarting the
recurrence relation with a new set of vectors once a certain number of
vectors have been generated. The optimization condition is maintained
over that particular set of vectors, however, Knoll and McHugh note
that this ultimately slows the convergence rate as compared to keeping
all vectors as the new set of vectors is not necessarily orthogonal to
the previous set, and therefore not optimal over the entire iteration
procedure. The orthogonality condition can be relaxed by using a
recurrence relation that does not generate a strictly orthonormal
basis for the Krylov subspace such as the Lanzcos biorthogonalization
procedure, resulting in memory savings due to the shorter Lanzcos
recurrence relation. Based on their numerical analysis, they observed
GMRES to be the most robust for any given nonlinear problem and
therefore used it for subsequent work in this area.

\subsection{Automatic Differentiation for Jacobian Generation}
\label{subsec:automatic_differentiation}
If it is acceptable to store the actual Jacobian matrix, methods are
available to construct it without requiring hand-coding and evaluating
derivatives, thus eliminating the associated issues. In addition, if
any additional equations are added to the system or a higher order
functional approximation is desired, it would be useful to avoid
regenerating and coding these derivatives. Becoming more prominent in
the 1990's, \textit{automatic differentiation} is a mechanism by which
the derivatives of a function can be generated automatically by
evaluating it. Automatic differentiation is built on the concept that
all functions discretely represented in a computer are ultimately
represented by elementary mathematical operations. If the chain rule
is applied to those elementary operations, then the derivatives of
those functions can be computed to the order of accuracy of their
original discretization in a completely automated way
\cite{averick_computing_1994}.

The work of Bartlett and others \cite{bartlett_automatic_2006}
extended initial Fortran-based work in the area of automatic
differentiation implementations to leverage the parametric type and
operator overloading features of C++ \cite{stroustrup_c++_1997}. They
formulate the differentiation problem from an element viewpoint by
assuming that a global Jacobian can be assembled from local element
function evaluations of $e_k : \mathbb{R}^{n_k} \rightarrow
\mathbb{R}^{m_k}$, similar to the finite element assembly procedure
as:
\begin{equation}
  \ve{J}(\ve{u}) = \sum_{i=1}^N \ve{Q}^T_i \ve{J}_k \ve{P}_i\:,
  \label{eq:fad_global_jacobian}
\end{equation}
where $\ve{J}_{k_i} = \partial e_{k_i} / \partial P_i u$ is the
$k^{th}$ element function Jacobian, $\ve{Q} \in \mathbb{R}^{n_{k_i}
  \times N}$ is a projector onto the element domain and $\ve{P} \in
\mathbb{R}^{m_{k_i} \times N}$ a projector onto the element range for
$\ve{F}(\ve{u}) \in \mathbb{R}^{N \times N}$. The Jacobian matrix for
each element will therefore have entirely local data in a dense
structure, eliminating the need for parallel communication and sparse
techniques during differentiation. Only when all local differentials
are computed does communication of the Jacobian occur through
gather/scatter operations in order to properly assemble it.

@@

%%---------------------------------------------------------------------------%%
\section{The FANM Method\ }
\label{sec:fanm}
In physics codes based on nonlinear equations sets, Newton-Krylov
methods are the primary means of generating a fully consistent
solution scheme
\cite{evans_development_2006,evans_enhanced_2007,gaston_parallel_2009,godoy_parallel_2012}. It
is common that for large scale simulations these problems are memory
limited due to the subspaces generated by robust Krylov methods which
may often build hundreds of vectors during a Newton step. Often, a
matrix-free approach is chosen to relax memory requirements over
directly generating the Jacobian matrix and to facilitate the
implementation. However, these matrix-free methods suffer from the
first order error introduced by the Jacobian approximation in poorly
scaled problems \citep{kelley_iterative_1995}. In addition, it has
been observed that the savings induced by the matrix-free approach is
eventually amortized over a number of nonlinear iterations where it
becomes more efficient computationally to instead form the Jacobian
\citep{knoll_newton-krylov_1995}.

@@

At first glance, FANM is simply a variant of an inexact Newton method
with a specific requirement of a fully formed Jacobian. However, FANM
has the potential to provide value in several areas. First, if
iterative or parallel performance of MCSA can be demonstrated to
better than that of GMRES or similar subspace methods for asymmetric
systems, FANM may outperform a Newton-Krylov method. Second, for
situations where memory is a constraint, especially in situations
where the Jacobian is generated often if not at every Newton
iteration, not building a subspace for ill-conditioned problems will
alleviate the associated cost. However, until the preconditioning
issues associated with MCSA as presented in the previous chapter using
the $SP_N$ equations can be resolved, this hypothesis can
unfortunately not be researched. Explicit algebraic preconditioning
used in conventional solution methods will almost certainly be
required to reduce the spectral radius of most Jacobian operators for
MCSA to apply and therefore the current preconditioning strategy will
destroy all potential memory benefits of FANM.

%%---------------------------------------------------------------------------%%
\section{Navier-Stokes Benchmark Problems\ }
\label{sec:ns_benchmarks}
To verify the FANM method for nonlinear problems, we choose benchmark
solutions for the 2-dimensional, steady, incompressible Navier-Stokes
equations on a rectilinear grid in much the same way as Shadid and
Pawlowski's work on Newton-Krylov methods for the solution of these
equations \cite{shadid_inexact_1997,pawlowski_globalization_2006}. We
define these equations as follows:
\begin{subequations}
  \begin{gather}
    \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
    \ve{g} = \ve{0}
    \label{eq:ns_momentum}\\
    \nabla \cdot \ve{u} = 0
    \label{eq:ns_continuity}\\
    \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0\:,
    \label{eq:ns_energy}
  \end{gather}
  \label{eq:navier_stokes}
\end{subequations}
where $\rho$ is the fluid density, $\ve{u}$ is the fluid velocity,
$C_p$ the specific heat capacity at constant pressure of the fluid,
$T$ the temperature of the fluid, and $\ve{g}$ the acceleration due to
gravity. Eq~(\ref{eq:ns_momentum}) provides momentum transport,
Eq~(\ref{eq:ns_continuity}) provides the mass balance, and
Eq~(\ref{eq:ns_energy}) provides energy transport with viscous
dissipation effects neglected. In addition, we close the system with
the following equations:
\begin{subequations}
  \begin{gather}
    \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
    \label{eq:ns_stress_tensor}\\
    \ve{q} = - k \nabla T\:,
    \label{eq:ns_heat_flux}
  \end{gather}
  \label{eq:ns_closure}
\end{subequations}
where $\ve{T}$ is the stress tensor, $P$ is the hydrodynamic pressure,
$\mu$ is the dynamic viscosity of the fluid, $\ve{q}$ is the heat flux
in the fluid, and $k$ is the thermal conductivity of the fluid. This
set of strongly coupled equations possesses both the nonlinearities
and asymmetries that we are seeking for qualification of the FANM
method. Further, physical parameters within these equations can be
tuned to enhance the nonlinearities. We will then apply these
equations to the following three standard benchmark problems that
cover convection dominated flow and flow dominated by inertial
forces in two different geometries.

@@

Isotherms of the temperature solution for these equations are given in
Figure~\ref{fig:convection_isotherms} for Rayleigh numbers of
\sn{1}{3}, \sn{1}{4}, \sn{1}{5} and \sn{1}{6}. Note that as the
Rayleigh number is increased, the fluid begins to rotate in the
clockwise direction. This causes the temperature gradients to rotate
as well with the fluid, given the increased rotation observed for the
isotherms in the figure. As the Rayleigh number increases, thermal
energy and momentum transport in the system become more dominated by
convection rather than diffusion, causing the nonlinearities in the
advective derivatives to grow.

@@

In this problem, the inflow condition is specified by a fully-formed
parabolic flow profile over a zero velocity boundary representing a
step. For this work, a slightly different parabolic formulation for
the inlet velocities was used than in the benchmark. In the benchmark
solutions, the average velocity in the parabola was 2/3 that of the
maximum velocity while the formula given in
Figure~\ref{fig:backward_facing_step} produces a more plug-like flow
with an average velocity 3/4 that of the maximum velocity. The flow
over this step will generate a recirculating backflow under the inlet
flow towards the step and at higher Reynolds numbers a recirculating
zone will form farther downstream from the step. As in the lid driven
cavity problem, the nonlinear behavior of this benchmark and the
difficulty in obtaining a solution is dictated by the Reynolds number
of the fluid.

@@

Results in this and the following section were produced by the Drekar
multiphysics code base being developed at Sandia National Laboratories
for fluid flow and other multiple physics simulations
\cite{pawlowski_drekar_2012}. Drekar is a massively parallel physics
framework that utilizes a finite element formulation of the
Navier-Stokes equations and Newton-Krylov methods with Jacobian
generation through automatic differentiation. To achieve these
results, a Monte Carlo solver generated for this work was injected
into the nonlinear solver sequence within Drekar to form an
implementation of the FANM method. For the Newton-Krylov solutions,
the same implementation of GMRES used for the $SP_N$ verification from
the Trilinos scientific computing library Aztec
\cite{heroux_overview_2005} was also used here as a means of
comparison and FANM benchmarking.

@@

A Drekar model of thermal convection cavity problem was solved using
both the Newton-Krylov solver and the FANM solver for Rayleigh numbers
of \sn{1}{3}, \sn{1}{4}, \sn{1}{5} and \sn{1}{6} on a $50 \times 50$
square grid. Table~\ref{tab:thermal_convection_parameters} gives the
physical parameters of the problem and
Table~\ref{tab:convection_mcsa_parameters} gives the MCSA solver
parameters used within the FANM method. Variation in the Rayleigh
number of the system was achieved by keeping the parameters given in
Table~\ref{tab:thermal_convection_parameters} constant and varying
temperature of the hot side of the cavity. In addition to multigrid
preconditioning for the linear solution at each Newton step,
backtracking as outlined in Shadid's work \cite{shadid_inexact_1997}
was used as a means of globalization to assist the Newton correction
computation while the forcing term at all Rayleigh numbers but
\sn{1}{6} was computed at each Newton step using method 2 from
Shadid's work. In this forcing term method, nonlinear residuals from
the two previous Newton steps are used to inform the choice of the
forcing term for the next step. At a Rayleigh number of \sn{1}{6}, a
constant forcing term of \sn{1}{-4} was use to give better convergence
properties.

@@

At each Rayleigh number the fluid velocity in the x and y directions,
fluid pressure, and fluid temperature values calculated with FANM were
verified by computing the point-wise absolute difference with the
Newton-Krylov results. No variation in solution was noted for each of
the computations larger than the floating point tolerance reported by
Drekar with an absolute difference value of zero observed at all
spatial locations for each quantity.

@@

At each Reynolds number the fluid velocity in the x and y directions
and fluid pressure values calculated with FANM were verified by
computing the point-wise absolute difference with the Newton-Krylov
results. No variation in solution was noted for each of the
computations larger than the floating point tolerance reported by
Drekar with an absolute difference value of zero observed at all
spatial locations for each quantity.

@@

As the Rayleigh number increases, the problem becomes more difficult
to solve and therefore a more aggressive preconditioner should
maintain iterative performance of the nonlinear method. This was
observed to be true for the \sn{1}{6} Rayleigh number case. For this
calculation, the number of multigrid cycle applications was bumped up
to 3 from 1 and FANM converged in the same number of nonlinear
iterations as the Newton-Krylov method.

@@

In addition to nonlinear iterative performance, performance of the
inner linear solves is also of interest. Here, we seek to maintain a
minimum number of total linear solver iterations over the course of
all nonlinear iterations in order to limit matrix-vector multiplies
and other costly
operations. Table~\ref{tab:convection_linear_iter_comparison} gives
the total number of GMRES iterations required to converge the problem
for the Newton-Krylov method and the total number of MCSA iterations
for the FANM method at each Rayleigh number used. For all cases, fewer
MCSA iterations were required than GMRES iterations to converge the
problem even with both solvers preconditioned with the same multigrid
package and the same parameters. For the \sn{1}{6} Rayleigh number
case where a constant forcing term of \sn{1}{-4} was kept, this
difference is most drastic with FANM requiring only 64\% of the linear
iterations required by the Newton-Krylov scheme with the linear solver
per nonlinear iteration given for this case in
Figure~\ref{fig:ra1e6_linear_iters}. Even the \sn{1}{5} Rayleigh
number case that required more Newton iterations required 5 fewer MCSA
iterations to converge than GMRES iterations. In all cases, MCSA
improved on the iterative performance of using only the Richardson
iteration through the residual Monte Carlo acceleration. If the performance of
the MCSA implementation used here were to be potentially optimized in
the future, this improved iterative performance could also mean
improved time to solution.

@@

Although the iterative performance of FANM is excellent for the
thermal convection cavity problem, the explicit preconditioning
strategy causes the same issues as was observed for the $SP_N$
equations in the previous
chapter. Table~\ref{tab:convection_speedup_comparison} gives the
speedup of the Newton-Krylov method over FANM. For all Rayleigh
numbers, we see a fairly constant speedup of O(400) when using
Newton-Krylov over FANM meaning that the implementation of the FANM
method used here is significantly slower. As the inexact Newton scheme
in Drekar was modified by simply swapping out the linear solver used
at each Newton step, this increased compute time is coming directly
from using MCSA as the linear solver and the overhead due to
preconditioning. When preconditioning overhead for the $SP_N$
equations was considered as shown by the data in
Figure~\ref{fig:spn_comparison_prec_time}, a similar speedup of
approximately two orders of magnitude was noted between the Krylov
methods and MCSA showing consistency between those results and the
convection cavity benchmark results presented. In addition, the fairly
constant speedup values as a function of the primary problem parameter
show that FANM has effectively the same time complexity as a
Newton-Krylov method and the implementation timings simply differ by a
constant value, also consistent with the $SP_N$ observations.

@@

>>>>

In this chapter the fundamentals of Newton's method will be presented
along with a brief discussion of inexact Newton methods. Newton-Krylov
methods, a variant of inexact Newton methods, will be outlined
including an automated technique for generating the Jacobian commonly
used in today's applications. Following this, a new nonlinear solution
algorithm utilizing Monte Carlo Synthetic Acceleration will be
presented and implementation details will be discussed. Using the new
algorithm and a production fluid physics code, a group of challenging
benchmark problems for the Navier-Stokes equations will be used to
verify the correctness of the method relative to solutions obtained
using conventional methods in various flow regimes and
geometries. Finally, we will compare the performance of the new method
against contemporary Newton-Krylov methods for the same benchmark
problems.

@@

\subsection{Inexact Newton Methods}
\label{subsec:inexact_newton_methods}
Inexact Newton methods arise when the Jacobian operator is not exactly
inverted as initially described by Dembo and others
\cite{dembo_inexact_1982}. For common sparse nonlinear systems, which
in turn yield a sparse Jacobian matrix, this situation occurs when
conventional iterative methods are applied. In their definition, Dembo
formulated inexact methods such that they are independent of the
linear method used to solve for the Newton correction and therefore
are amenable to use with any linear solver. Furthermore, they bind the
convergence of the outer nonlinear iteration to the inner linear
iteration such that:
\begin{equation}
  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
  ||\ve{F}(\ve{u}^k)||\:,
  \label{eq:inexact_newton_forcing}
\end{equation}
where $\eta^k \in [0,1)$ is defined as the \textit{forcing term} at
  the $k$ iterate. Eq~(\ref{eq:inexact_newton_forcing}) then states
  that the residual generated by the linear solver is bound by the
  nonlinear residual and how tightly it is bound is defined by the
  forcing term. This is useful in that we can vary how tightly coupled
  the convergence of the linear iterations used to generate the Newton
  correction is to the nonlinear iteration by relaxing or tightening
  the convergence properties on the linear iterative method. As a
  result, strategies for determining the forcing term can vary
  depending on the problem type and can greatly affect the convergence
  of the method or even prohibit convergence
  \cite{eisenstat_choosing_1996}. In addition, \textit{globalization
    methods} may be used to modify the Newton correction in a more
  desirable direction such that nonlinear convergence properties can
  be improved when the initial guess for $\ve{u}$ is poor
  \cite{pawlowski_globalization_2006}.

\subsection{Newton-Krylov Methods\ }
\label{subsec:newton_krylov_methods}
A form of inexact Newton methods, \textit{Newton-Krylov methods} are
nonlinear iterative methods that leverage a Krylov subspace method as
the linear solver for generating the Newton correction
\cite{kelley_iterative_1995}. As outlined in
Appendix~\ref{ch:linear_problem}, Krylov methods are robust and enjoy
efficient parallel implementations on modern
architectures. Furthermore, their lack of explicit dependence on the
operator make them easier to implement than other methods. In most
nonlinear problems, the Jacobian operator is generally non-symmetric
and therefore either Krylov methods with long recurrence relations
that can handle non-symmetric systems must be considered or the Newton
correction system must be preconditioned such that the operator is
symmetric and short recurrence relation methods can be potentially be
used.

With many Krylov methods available, which to use with the Newton
method is dependent on many factors including convergence rates and
memory usage. Several studies have been performed to investigate this
\cite{mchugh_inexact_1993,knoll_newton-krylov_1995}. In their
numerical studies in 1995, Knoll and McHugh used the set of highly
nonlinear and stiff convection-diffusion-reaction equations to solve a
set of tokamak plasma problems with the goal of measuring solver
performance with Newton's method. Based on their numerical analysis,
they observed GMRES to be the most robust for any given nonlinear
problem and therefore used it for subsequent work in this area. Using
their analysis, this work will also use GMRES as the linear solver for
all Newton-Krylov reference calculations.

\subsection{Automatic Differentiation for Jacobian Generation}
\label{subsec:automatic_differentiation}
If it is acceptable to store the actual Jacobian matrix, methods are
available to construct it without requiring hand-coding and evaluating
derivatives, thus eliminating the associated issues. In addition, if
any additional equations are added to the system or a higher order
functional approximation is desired, it would be useful to avoid
regenerating and coding these derivatives. Becoming more prominent in
the 1990's, \textit{automatic differentiation} is a mechanism by which
the derivatives of a function can be generated automatically by
evaluating it. Automatic differentiation is built on the concept that
all functions discretely represented in a computer are ultimately
formed by elementary mathematical operations. If the chain rule is
applied to those elementary operations, then the derivatives of those
functions can be computed to the order of accuracy of their original
discretization in a completely automated way
\cite{averick_computing_1994}.

The work of Bartlett and others \cite{bartlett_automatic_2006}
extended initial Fortran-based work in the area of automatic
differentiation implementations to leverage the parametric type and
operator overloading features of C++ \cite{stroustrup_c++_1997}. They
formulate the differentiation problem from an element viewpoint by
assuming that a global Jacobian can be assembled from local element
function evaluations of $e_k : \mathbb{R}^{n_k} \rightarrow
\mathbb{R}^{m_k}$, similar to the finite element assembly procedure
as:
\begin{equation}
  \ve{J}(\ve{u}) = \sum_{i=1}^N \ve{Q}^T_i \ve{J}_k \ve{P}_i\:,
  \label{eq:fad_global_jacobian}
\end{equation}
where $\ve{J}_{k_i} = \partial e_{k_i} / \partial P_i u$ is the
$k^{th}$ element function Jacobian, $\ve{Q} \in \mathbb{R}^{N \times
  n_{k_i}}$ is a projector onto the element domain and $\ve{P} \in
\mathbb{R}^{m_{k_i} \times N}$ a projector from the element range for
$\ve{F}(\ve{u}) \in \mathbb{R}^{N \times N}$. The Jacobian matrix for
each element will therefore have entirely local data in a dense
structure, eliminating the need for parallel communication and sparse
techniques during differentiation. Only when all local differentials
are computed does communication of the Jacobian occur through
gather/scatter operations in order to properly assemble it.

@@

%%---------------------------------------------------------------------------%%
\section{The FANM Method\ }
\label{sec:fanm}
In physics codes based on nonlinear equations sets, Newton-Krylov
methods are the primary means of generating a fully implicit
solution scheme
\cite{evans_development_2006,evans_enhanced_2007,gaston_parallel_2009,godoy_parallel_2012}. It
is common that for large scale simulations these problems are memory
limited due to the subspaces generated by robust Krylov methods which
may often build hundreds of vectors during a Newton step. Often, a
matrix-free approach is chosen to relax memory requirements over
directly generating the Jacobian matrix and to facilitate the
implementation. However, these matrix-free methods suffer from the
first order error introduced by the Jacobian approximation in poorly
scaled problems \citep{kelley_iterative_1995}. In addition, it has
been observed that the savings induced by the matrix-free approach is
eventually amortized over a number of nonlinear iterations where it
becomes more efficient computationally to instead form the Jacobian
\citep{knoll_newton-krylov_1995}.

@@

At first glance, FANM is simply a variant of an inexact Newton method
with a specific requirement of a fully formed Jacobian. However, FANM
has the potential to provide value in several areas. First, if
iterative or parallel performance of MCSA can be demonstrated to
better than that of GMRES or similar subspace methods for asymmetric
systems, FANM may outperform a Newton-Krylov method. Second, for
situations where memory is a constraint, especially in situations
where the Jacobian is generated often if not at every Newton
iteration, not building a subspace for ill-conditioned problems will
alleviate the associated cost. However, until the preconditioning
issues associated with MCSA as presented in the previous chapter using
the $SP_N$ equations can be resolved, this hypothesis can
unfortunately not be researched. Explicit algebraic preconditioning
used in conventional solution methods will almost certainly be
required to reduce the spectral radius of most Jacobian operators for
MCSA to apply and therefore the current preconditioning strategy will
eliminate all potential memory benefits of FANM.

%%---------------------------------------------------------------------------%%
\section{Navier-Stokes Benchmark Problems\ }
\label{sec:ns_benchmarks}
To verify the FANM method for nonlinear problems, we choose benchmark
solutions for the 2-dimensional, steady, incompressible Navier-Stokes
equations on a rectilinear grid as outlined in Shadid and Pawlowski's
work on Newton-Krylov methods
\cite{shadid_inexact_1997,pawlowski_globalization_2006}. We define
these equations as follows:
\begin{subequations}
  \begin{gather}
    \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
    \ve{g} = \ve{0}
    \label{eq:ns_momentum}\\
    \nabla \cdot \ve{u} = 0
    \label{eq:ns_continuity}\\
    \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0\:,
    \label{eq:ns_energy}
  \end{gather}
  \label{eq:navier_stokes}
\end{subequations}
where $\rho$ is the fluid density, $\ve{u}$ is the fluid velocity,
$C_p$ the specific heat capacity at constant pressure of the fluid,
$T$ the temperature of the fluid, and $\ve{g}$ the acceleration due to
gravity. Eq~(\ref{eq:ns_momentum}) provides momentum transport,
Eq~(\ref{eq:ns_continuity}) provides the mass balance, and
Eq~(\ref{eq:ns_energy}) provides energy transport with viscous
dissipation effects neglected. In addition, we close the system with
the following equations:
\begin{subequations}
  \begin{gather}
    \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
    \label{eq:ns_stress_tensor}\\
    \ve{q} = - k \nabla T\:,
    \label{eq:ns_heat_flux}
  \end{gather}
  \label{eq:ns_closure}
\end{subequations}
where $\ve{T}$ is the stress tensor, $P$ is the hydrodynamic pressure,
$\mu$ is the dynamic viscosity of the fluid, $\ve{q}$ is the heat flux
in the fluid, and $k$ is the thermal conductivity of the fluid. This
set of strongly coupled equations possesses both the nonlinearities
and asymmetries that we are seeking for qualification of the FANM
method. Further, physical parameters within these equations can be
tuned to enhance the nonlinearities. We next apply these equations to
the following three standard benchmark problems that cover convection
dominated flow and flow dominated by inertial forces in two different
geometries.

@@

Isotherms of the temperature solution for these equations are given in
Figure~\ref{fig:convection_isotherms} for Rayleigh numbers of
\sn{1}{3}, \sn{1}{4}, \sn{1}{5} and \sn{1}{6}. Note that as the
Rayleigh number is increased, the fluid begins to rotate in the
clockwise direction. This causes the temperature gradients to rotate
as well with the fluid, given the increased rotation observed for the
isotherms in the figure. As the Rayleigh number and rotation
increases, thermal energy and momentum transport in the system become
more dominated by convection rather than diffusion, causing the
nonlinearities in the advective derivatives to grow.

@@

In this problem, the inflow condition is specified by a fully-formed
parabolic flow profile over a zero velocity boundary representing a
step. For this work, a slightly different parabolic formulation for
the inlet velocities was used than in the benchmark. In the benchmark
solutions, the average velocity in the parabola was 2/3 that of the
maximum velocity while the formula given in
Figure~\ref{fig:backward_facing_step} produces a more plug-like flow
with an average velocity 3/4 that of the maximum velocity. The flow
over this step will generate a recirculating backflow under the inlet
flow towards the step and at higher Reynolds numbers a recirculation
zone will form farther downstream from the step. As in the lid driven
cavity problem, the nonlinear behavior of this benchmark and the
difficulty in obtaining a solution is dictated by the Reynolds number
of the fluid.

@@

Results in this and the following section were produced by the Drekar
multiphysics code base being developed at Sandia National Laboratories
for fluid flow and other multiple physics simulations
\cite{pawlowski_drekar_2012}. Drekar is a massively parallel physics
framework that utilizes a finite element formulation of the
Navier-Stokes equations and Newton-Krylov methods with Jacobian
generation through automatic differentiation. To achieve the results
presented here, a Monte Carlo solver generated for this work was
injected into the nonlinear solver sequence within Drekar to form an
implementation of the FANM method. For the Newton-Krylov solutions,
the same implementation of GMRES used for the $SP_N$ verification from
the Trilinos scientific computing library Aztec
\cite{heroux_overview_2005} was also used here as a means of
comparison and FANM benchmarking.

@@

A Drekar model of thermal convection cavity problem was solved using
both the Newton-Krylov solver and the FANM solver for Rayleigh numbers
of \sn{1}{3}, \sn{1}{4}, \sn{1}{5} and \sn{1}{6} on a $50 \times 50$
square grid. Table~\ref{tab:thermal_convection_parameters} gives the
physical parameters of the problem and
Table~\ref{tab:convection_mcsa_parameters} gives the MCSA solver
parameters used within the FANM method. Variation in the Rayleigh
number of the system was achieved by keeping the parameters given in
Table~\ref{tab:thermal_convection_parameters} constant and varying the
temperature of the hot side of the cavity. In addition to multigrid
preconditioning for the linear solution at each Newton step,
backtracking as outlined in Shadid's work \cite{shadid_inexact_1997}
was used as a means of globalization to assist the Newton correction
computation while the forcing term at all Rayleigh numbers but
\sn{1}{6} was computed at each Newton step using method 2 from
Shadid's work. In this forcing term method, nonlinear residuals from
the two previous Newton steps are used to inform the choice of the
forcing term for the next step. At a Rayleigh number of \sn{1}{6}, a
constant forcing term of \sn{1}{-4} was use to give better convergence
properties.

@@

At each Rayleigh number the fluid velocity in the x and y directions,
fluid pressure, and fluid temperature values calculated with FANM were
verified by computing the point-wise absolute difference with the
Newton-Krylov results. No variation in the solution was noted for each
of the computations larger than the floating point tolerance reported
by Drekar with an absolute difference value of zero computed at all
spatial locations for each physical quantity.

@@

At each Reynolds number the fluid velocity in the x and y directions
and fluid pressure values calculated with FANM were verified by
computing the point-wise absolute difference with the Newton-Krylov
results. No variation in the solution was noted for each of the
computations larger than the floating point tolerance reported by
Drekar with an absolute difference value of zero computed at all
spatial locations for each physical quantity.

@@

As the Rayleigh number increases, the problem becomes more difficult
to solve and therefore a more aggressive preconditioner should
maintain iterative performance of the nonlinear method. This was
observed to be true for the \sn{1}{6} Rayleigh number case. For this
calculation, the number of multigrid cycle applications was increased
to 3 from 1 and FANM converged in the same number of nonlinear
iterations as the Newton-Krylov method.

@@

In addition to nonlinear iterative performance, performance of the
inner linear solves is also of interest. Here, we seek to maintain a
minimum number of total linear solver iterations over the course of
all nonlinear iterations in order to limit matrix-vector multiplies
and other costly
operations. Table~\ref{tab:convection_linear_iter_comparison} gives
the total number of GMRES, FANM, and Richardson iterations required to
converge the Newton method at each Rayleigh number used. For all
cases, fewer MCSA iterations were required than GMRES iterations to
converge the problem even with both solvers preconditioned with the
same multigrid package and the same parameters. For the \sn{1}{6}
Rayleigh number case where a constant forcing term of \sn{1}{-4} was
kept, this difference is most drastic with FANM requiring only 64\% of
the linear iterations required by the Newton-Krylov scheme with the
linear solver per nonlinear iteration given for this case in
Figure~\ref{fig:ra1e6_linear_iters}. Even the \sn{1}{5} Rayleigh
number case that required more Newton iterations required 5 fewer MCSA
iterations to converge than GMRES iterations. In all cases, MCSA
improved on the iterative performance of using only the Richardson
iteration through the residual Monte Carlo acceleration. If the
performance of the MCSA implementation used here were to be
potentially optimized in the future, this improved iterative
performance could also mean improved time to solution.

@@

Although the iterative performance of FANM is excellent for the
thermal convection cavity problem, the explicit preconditioning
strategy causes the same issues as was observed for the $SP_N$
equations in the previous
chapter. Table~\ref{tab:convection_speedup_comparison} gives the
speedup of the Newton-Krylov method over FANM. For all Rayleigh
numbers, we see a fairly constant speedup of O(400) when using
Newton-Krylov over FANM meaning that the implementation of the FANM
method used here is significantly slower.

@@

As the inexact Newton scheme in Drekar was modified by simply swapping
out the linear solver used at each Newton step, this increased compute
time is coming directly from using MCSA as the linear solver and the
overhead due to preconditioning. When preconditioning overhead for the
$SP_N$ equations was considered as shown by the data in
Figure~\ref{fig:spn_comparison_prec_time}, a similar speedup of
approximately two orders of magnitude was noted between the Krylov
methods and MCSA showing consistency between those results and the
convection cavity benchmark results presented. In addition, the fairly
constant speedup values as a function of the primary problem parameter
show that FANM has effectively the same time complexity as a
Newton-Krylov method and the implementation timings simply differ by a
constant value, also consistent with the $SP_N$ observations.
